---
title: "Measurement checks -- HPT (Czech data)"
subtitle: "DIF & multi-group CFA by ideological attitude (ideological contamination tests)"
author: "HPT and Extremism project"
date: "`r format(Sys.Date())`"
always_allow_html: true
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: true
    toc: true
    toc_depth: 3
    keep_tex: false
    
  md_document:
    variant: markdown
    preserve_yaml: false
    toc: false

# global chunk options
knitr:
  opts_chunk:
    dev: "png"          # ensures image files are written
    fig.path: "figures/" # directory for external images
    dpi: 300

fontsize: 10pt
geometry:
  - landscape
  - margin=0.7in

header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \setlength{\LTcapwidth}{\textwidth}
---

# Purpose & plan

We assess whether *ideological attitude groups* respond differently to the HPT items **even at equal underlying HPT competence**. Concretely:

* Create **low/high ideology** groups from FR-LF-mini and KSA-3 (see codebook).  
* Run **joint omnibus GRM DIF** tests for each HPT item.
* Run **multi-group CFA** (configural → metric → scalar).
* The registered H4 concerned positive ideology-related DIF on CONT items,
  evaluated with a continuous-ideology MIMIC analysis. The current joint
  omnibus GRM DIF and MG-CFA analyses are post-registration and do not
  implement the registered continuous-ideology MIMIC analysis. Evidence of DIF
  or scalar non-invariance would indicate measurement noninvariance between the
  extreme ideology groups, limiting score comparability without itself
  establishing a directional contamination mechanism.

# Setup

```{r,setup, message=FALSE, warning=FALSE}
options(width = 120)
library(dplyr)
library(tidyr)
library(ggplot2)
library(psych)
library(knitr)
library(stringr)
library(janitor)
library(difR)        # DIF for ordinal items
library(lavaan)      # CFA / invariance
library(semTools)    # helpers
library(car)         # recode
library(mirt)
source("submissions/pci_psychology/scoring_helpers.R")
```

# Data

```{r,load-data}
# Load the dataset created in 00_data-preparation
load("normalised_responses.RData")
stopifnot(exists("normalised_responses"))
dat_raw <- normalised_responses

# Cluster vars
dat_raw <- dat_raw %>%
  mutate(
    school_id   = as.factor(school_id),
    class_label = as.factor(class_label),
    class_id    = interaction(school_id, class_label, drop = TRUE)
  )

# Coerce HPT items to numeric early (1-4 expected in codebook)
hpt_items <- c(paste0("POP",1:3), paste0("ROA",1:3), paste0("CONT",1:3))
dat_raw <- dat_raw %>% mutate(across(all_of(hpt_items), ~ suppressWarnings(as.numeric(.))))

# Reverse POP item-wise (so higher = more contextualised)
POP_rev_items <- paste0("POP", 1:3)
dat_raw <- dat_raw %>%
  mutate(across(all_of(POP_rev_items), ~ 5 - ., .names = "{.col}_rev")) %>%
  mutate(
    HPT_POP_raw = scale_mean(., POP_rev_items, min_answered = 2),
    HPT_POP_rev = scale_mean(., paste0(POP_rev_items, "_rev"), min_answered = 2),
    HPT_CONT    = scale_mean(., paste0("CONT", 1:3), min_answered = 2),
    HPT_ROA     = scale_mean(., paste0("ROA", 1:3), min_answered = 2),

    # Canonical composites (CTX6 is our stable default)
    HPT_CTX6 = rowMeans(cbind(HPT_POP_rev, HPT_CONT), na.rm = FALSE),
    HPT_TOT9 = rowMeans(cbind(HPT_POP_rev, HPT_CONT, HPT_ROA),
                        na.rm = FALSE)
  )
```

We use variables as defined in the **codebook** (metadata; KN1-KN6; POP1-POP3; ROA1-ROA3; CONT1-CONT3; FR-LF mini RD1-RD3 & NS1-NS3; KSA-3 A1-A3, U1-U3, K1-K3; SDR1-SDR5). 

## Scoring & grouping

* **HPT items** are 1-4. We reverse only *presentism* items (`POP1-POP3`: `5 - POP*`) so that a single higher-is-better direction is used for scale construction and MG-CFA. DIF uses original item codings (reversal is not required for DIF detection).
* **Ideology composite**: FR-LF-mini (RD1-3 + NS1-3) and KSA-3 (9 items). We z-score the two scale means and average them → **IDEO_Z**. *Low* = bottom 33%, *High* = top 33% (middle third excluded to sharpen contrasts). 
* **Controls**: prior knowledge (sum KN1-KN6), social desirability (SDR1-SDR5; note SDR2-SDR4 are already reversed upstream per codebook).

```{r,scoring, message=FALSE}
# Select blocks
frlf_items <- c(paste0("RD",1:3), paste0("NS",1:3))
ksa_items  <- c(paste0("A",1:3), paste0("U",1:3), paste0("K",1:3))
kn_items   <- paste0("KN",1:6)
sdr_items  <- paste0("SDR",1:5)

# Coerce predictors to numeric
num_blocks <- c(frlf_items, ksa_items, kn_items, sdr_items)

dat <- dat_raw %>%
  mutate(across(all_of(num_blocks), ~ suppressWarnings(as.numeric(.)))) %>%
  # Scale scores
  mutate(
    HPT_total = HPT_TOT9,
    HPT_POPR  = HPT_POP_rev,
    FRLF_mean = scale_mean(., frlf_items, min_answered = 4),
    KSA_mean  = scale_mean(., ksa_items, min_answered = 7),
    KN_sum    = rowSums(across(all_of(kn_items)), na.rm = TRUE),
    SDR_mean  = scale_mean(., sdr_items, min_answered = 4)
  ) %>%
  mutate(
    FRLF_z = as.numeric(scale(FRLF_mean)),
    KSA_z  = as.numeric(scale(KSA_mean)),
    IDEO_Z = (FRLF_z + KSA_z) / 2
  )

# Define tertile groups
qs <- quantile(dat$IDEO_Z, probs = c(.3334, .6666), na.rm = TRUE)
dat <- dat %>%
  mutate(
    ideology_group = case_when(
      IDEO_Z <= qs[1] ~ "Low",
      IDEO_Z >= qs[2] ~ "High",
      !is.na(IDEO_Z) ~ "Mid"
    )
  )

kable(dat %>% count(ideology_group), caption = "Group sizes (Low/High ideology; Mid excluded from group-wise tests)")
```

> **Note.** We focus on *Low* vs *High* to maximise contrast for DIF/MG-CFA. *Mid* is retained for descriptives only.

# Descriptives (checks)

```{r,desc}
desc_tbl <- dat %>%
  group_by(ideology_group) %>%
  summarise(n = n(),
            HPT_total = mean(HPT_total, na.rm = TRUE),
            KN_sum    = mean(KN_sum,    na.rm = TRUE),
            SDR_mean  = mean(SDR_mean,  na.rm = TRUE)) %>%
  arrange(match(ideology_group, c("Low","Mid","High")))
kable(desc_tbl, digits = 2, caption = "Descriptives by ideology group (means)")
```

# DIF analysis (ordinal, item-by-item)

**Goal.** Do Low/High ideology groups differ on individual HPT items in a
joint omnibus likelihood-ratio DIF test? Each test jointly releases that item's
slope and threshold constraints. Raw *p*-values are Bonferroni-adjusted across
the nine items and evaluated at familywise alpha = .05 (equivalent per-test
alpha = .0056).

```{r,DIF-prep, message=FALSE}
## Keep only Low/High groups
anal <- dat[dat$ideology_group %in% c("Low","High"), , drop = FALSE]

## HPT item list in original coding
hpt_items <- c("POP1","POP2","POP3","ROA1","ROA2","ROA3","CONT1","CONT2","CONT3")
stopifnot(all(hpt_items %in% names(anal)))

## Build item matrix
hpt_mat <- anal[, hpt_items, drop = FALSE]
for (j in seq_along(hpt_items)) hpt_mat[[j]] <- suppressWarnings(as.numeric(hpt_mat[[j]]))

## Group factor
grp <- factor(anal$ideology_group, levels = c("Low","High"))

## Drop rows with <2 answered items
keep <- rowSums(!is.na(hpt_mat)) >= 2
hpt_mat <- hpt_mat[keep, , drop = FALSE]
grp     <- droplevels(grp[keep])

stopifnot(nrow(hpt_mat) == length(grp), nlevels(grp) == 2)
print(table(grp))
```

```{r,DIF-run, message=FALSE, warning=FALSE}
# Constrained multi-group graded model, then DIF with scheme="drop"
mod_base <- multipleGroup(
  data       = hpt_mat,
  model      = 1,
  group      = grp,
  itemtype   = "graded",
  invariance = c("slopes", "intercepts", "free_means", "free_var")
)

params_all <- mirt::mod2values(mod_base)$name
unique_pars <- unique(params_all)
pars_to_test <- c(
  grep("^a", unique_pars, value = TRUE),
  grep("^d\\d+$", unique_pars, value = TRUE)
)
stopifnot(length(pars_to_test) > 0)

dif_out <- DIF(
  mod_base,
  which.par   = pars_to_test,
  scheme      = "drop",
  items2test  = colnames(hpt_mat),
  p.adjust    = "bonferroni",
  verbose     = FALSE
)

res_tbl <- as.data.frame(dif_out) %>%
  tibble::rownames_to_column("Item") %>%
  transmute(
    Item,
    X2,
    df,
    p,
    adj_p,
    Flag = ifelse(adj_p < .05, "YES", "no")
  )

kable(res_tbl, digits = 3,
      caption = "DIF omnibus item tests (mirt; graded). Each test jointly releases slope and threshold constraints. Raw p-values and Bonferroni-adjusted p-values are shown; familywise alpha = .05 (per-test alpha = .0056).")
```

```{r,DIF-plot, fig.width=9, fig.height=4, message=FALSE, warning=FALSE}
flagged <- res_tbl$Item[res_tbl$Flag == "YES"]
if (length(flagged) > 0) {
  which_item <- which(colnames(hpt_mat) == flagged[1])
  plot(mod_base, type = "trace", which.items = which_item,
       facet_items = FALSE, groups = levels(grp))
} else {
  plot.new(); text(0.5, 0.5,
                   "No omnibus DIF item was flagged after Bonferroni adjustment (familywise alpha = .05).")
}
```

```{r,table-s1-parameters}
grm_coefficients <- coef(mod_base, IRTpars = FALSE, simplify = TRUE)
stopifnot(isTRUE(all.equal(
  grm_coefficients$Low$items,
  grm_coefficients$High$items,
  check.attributes = FALSE
)))

table_s1_for_group <- function(group) {
  as.data.frame(grm_coefficients[[group]]$items) %>%
    tibble::rownames_to_column("Item") %>%
    transmute(Item, a1, d1, d2, d3)
}

table_s1 <- bind_rows(
  High = table_s1_for_group("High"),
  Low = table_s1_for_group("Low"),
  .id = "Group"
) %>%
  arrange(Item, factor(Group, levels = c("High", "Low"))) %>%
  mutate(across(c(a1, d1, d2, d3), ~ round(.x, 3)))

write.csv(table_s1, "outputs/table_s1_irt_parameters.csv", row.names = FALSE)
kable(filter(table_s1, Group == "High") %>% select(-Group), digits = 3,
      caption = "Table S1 source: constrained multi-group GRM item parameters from mod_base. Parameters are equal across Low and High ideology groups.")
```

# Multi-group CFA (configural → metric → scalar)

**Model.** We specify a **three-factor model** (POP_rev, ROA, CONT) with POP items reversed so that all loadings point to *more HPT-congruent* responses. We then test invariance across Low vs High ideology groups.

```{r,mgcfa-data}
# Build analysis frame with reversed POP and intact ROA/CONT
cfad <- dat %>%
  filter(ideology_group %in% c("Low","High")) %>%
  transmute(
    ideology_group, class_id,         # keep cluster id for reference (not used by lavaan here)
    POP1 = 5 - POP1,
    POP2 = 5 - POP2,
    POP3 = 5 - POP3,
    ROA1, ROA2, ROA3,
    CONT1, CONT2, CONT3
  )

# Coerce all item columns to numeric and enforce ordinal 1:4 range; replace out-of-range with NA
ord_items <- setdiff(names(cfad), c("ideology_group", "class_id"))
cfad <- cfad %>% mutate(across(all_of(ord_items), ~ suppressWarnings(as.numeric(.))))
cfad <- cfad %>% mutate(across(all_of(ord_items), ~ ifelse(. %in% 1:4, ., NA_real_)))

# If any columns had non 1-4 values, warn instead of stopping
bad_cols <- vapply(cfad[ord_items], function(x) any(!is.na(x) & !(x %in% 1:4)), logical(1))
if (any(bad_cols)) {
  warning(sprintf("Non-1:4 values were set to NA in: %s", paste(names(bad_cols)[bad_cols], collapse=", ")))
}
```

```{r,mgcfa-model}
model_3f <- '
  POP =~ POP1 + POP2 + POP3
  ROA =~ ROA1 + ROA2 + ROA3
  CONT =~ CONT1 + CONT2 + CONT3
'
```

```{r,mgcfa-fit, message=FALSE, warning=FALSE}
# Run invariance ladder with WLSMV on ordered items; DO NOT pass cluster (not supported with ordered)
fit_conf <- cfa(model_3f, data = cfad, group = "ideology_group",
                ordered = ord_items, estimator = "WLSMV")

fit_metr <- cfa(model_3f, data = cfad, group = "ideology_group",
                ordered = ord_items, estimator = "WLSMV",
                group.equal = c("loadings"))

fit_scal <- cfa(model_3f, data = cfad, group = "ideology_group",
                ordered = ord_items, estimator = "WLSMV",
                group.equal = c("loadings", "thresholds"))

get_fit <- function(fit) {
  fitMeasures(fit, c("chisq.scaled", "df.scaled", "pvalue.scaled",
                    "cfi.scaled", "rmsea.scaled", "srmr"))
}

fits <- bind_rows(
  Configural = get_fit(fit_conf),
  Metric     = get_fit(fit_metr),
  Scalar     = get_fit(fit_scal)
) %>%
  mutate(Model = c("Configural", "Metric", "Scalar")) %>%
  select(Model, everything())

fits %>% mutate(across(where(is.numeric), round, 3)) %>%
  kable(caption = "MG-CFA fit indices by invariance level (WLSMV).")
```

```{r,mgcfa-deltas}
deltas <- tibble(
  step  = c("Configural -> Metric", "Metric -> Scalar"),
  dCFI  = c(fits$cfi.scaled[2] - fits$cfi.scaled[1], fits$cfi.scaled[3] - fits$cfi.scaled[2]),
  dRMSEA= c(fits$rmsea.scaled[2] - fits$rmsea.scaled[1], fits$rmsea.scaled[3] - fits$rmsea.scaled[2])
)

deltas %>% mutate(across(where(is.numeric), round, 3)) %>%
  kable(caption = "Delta fit (CFI, RMSEA) across steps.")
```

> **How to read this.** If **metric holds** (small ΔCFI/ΔRMSEA), loadings are
> equivalent. If **scalar fails**, thresholds differ and group-mean comparisons
> are measurement-noninvariant; this warrants cautious interpretation but does
> not itself establish a directional ideological effect.

# (Optional) Two-factor robustness check

```{r,mgcfa-2f, eval=FALSE}
model_2f <- '
  CTX =~ POP1 + POP2 + POP3 + CONT1 + CONT2 + CONT3
  ROA =~ ROA1 + ROA2 + ROA3
'
measurementInvariance(model_2f, data = cfad, group = "ideology_group",
                      estimator = "WLSMV",
                      ordered = ord_items)
```

# Interpretation & reporting

## DIF summary

```{r,interp-dif}
kable(res_tbl, caption = "DIF results (for reference in text).")
```

## MG-CFA summary

```{r,interp-mgcfa}
kable(fits %>% mutate(across(where(is.numeric), round, 3)),
      caption = "MG-CFA fit to reference in text.")

kable(deltas %>% mutate(across(where(is.numeric), round, 3)), 
      caption = "Delta fit (CFI, RMSEA) thresholds.")
```

# Reproducibility appendix

```{r,session-info}
sessionInfo()
```
