---
title: "Supplement table and figure inventory"
author: "HPT and Extremism project"
date: "`r format(Sys.Date())`"
output:
  pdf_document:
    latex_engine: xelatex
  md_document:
    variant: markdown
---

# Purpose

This documentation report inventories the current supplementary tables and
figures and identifies their reproducible source files. It does not analyse
confidential source data.

# Supplement tables

```{r supplement-tables, echo=FALSE}
tables <- data.frame(
  Element = c("Table S1", "Table S2", "Table S3 and S3b", "Table S4",
              "Table S4b", "Table S5"),
  Source = c(
    "04_dif-and-mg-cfa-hpt-bias.Rmd -> outputs/table_s1_irt_parameters.csv",
    "01_measurement-checks.Rmd",
    "submissions/pci_psychology/revision_reporting_analyses.R",
    "submissions/pci_psychology/supplementary_materials.md",
    "submissions/pci_psychology/revision_reporting_analyses.R",
    "submissions/pci_psychology/supplementary_materials.md"
  ),
  stringsAsFactors = FALSE
)
knitr::kable(tables, caption = "Current supplementary-table inventory")
```

# Figures and revision outputs

```{r figures, echo=FALSE}
figures <- data.frame(
  Element = c("Revision relationship figure", "Measurement invariance and DIF",
              "Score distributions", "Coefficient plot", "Marginal effects"),
  Source = c(
    "submissions/pci_psychology/revision_reporting_analyses.R -> revision_outputs/",
    "osf_storage/scripts/fig02_measurement_invariance_and_dif.R",
    "osf_storage/scripts/fig03_score_distributions.R",
    "osf_storage/scripts/fig04_coefficient_plot.R",
    "osf_storage/scripts/fig05_marginal_effects.R"
  ),
  stringsAsFactors = FALSE
)
knitr::kable(figures, caption = "Current figure and revision-output inventory")
```

# Scope note

The markdown and PDF supplement are the presentation source for Tables S1-S5.
This report is an inventory, not a second presentation of the estimates.
