---
title: "Reproducibility report"
author: "HPT and Extremism project"
date: "`r format(Sys.Date())`"
output:
  pdf_document:
    latex_engine: xelatex
  md_document:
    variant: markdown
---

# Scope

This documentation report records reproducibility metadata for the public,
de-identified package. It does not load confidential data.

# Public-data checksums

```{r checksums, echo=FALSE}
data_files <- c("normalised_responses.RDS", "normalised_responses.xlsx")
checksum_command <- Sys.which("sha256sum")
if (!nzchar(checksum_command)) stop("sha256sum is required to record checksums.")
checksums <- system2(checksum_command, data_files, stdout = TRUE)
knitr::kable(data.frame(SHA256 = checksums),
             caption = "SHA-256 checksums for public analysis data")
```

# Seed policy

The numbered analytic reports set a seed only where stochastic processing is
used. Report 03 sets `1234`; the revision bootstrap uses `20260910`. Other
analyses are deterministic conditional on the declared data and package
versions. Exploratory scripts retain their own documented seeds.

# Declared package versions

```{r package-versions, echo=FALSE}
declared_packages <- c(
  "lavaan", "lme4", "lmerTest", "mirt", "psych", "semTools",
  "tidyverse", "janitor", "knitr", "rmarkdown"
)
package_versions <- vapply(declared_packages, function(package) {
  if (requireNamespace(package, quietly = TRUE)) {
    as.character(utils::packageVersion(package))
  } else {
    "not installed"
  }
}, character(1))
knitr::kable(data.frame(Package = declared_packages, Version = package_versions),
             caption = "Declared package versions available at render time")
```

# File map

```{r file-map, echo=FALSE}
file_map <- data.frame(
  File = c(
    "01_measurement-checks.Rmd", "02_descriptives-and-zero-order.Rmd",
    "03_multilevel-models-hypothesis-tests.Rmd", "04_dif-and-mg-cfa-hpt-bias.Rmd",
    "05_sensitivity-analyses.Rmd", "06_appendix-tables-and-figures.Rmd",
    "07_reproducibility-report.Rmd"
  ),
  Role = c(
    "Measurement checks", "Descriptives, correlations, and ICC components",
    "Multilevel focal models", "DIF and MG-CFA", "Sensitivity analyses",
    "Supplement inventory", "Reproducibility metadata"
  ),
  stringsAsFactors = FALSE
)
knitr::kable(file_map, caption = "Numbered-report file map")
```

# Session information

```{r session-info}
sessionInfo()
```
