\documentclass[12pt]{article}
\usepackage[margin=2.5cm]{geometry}
\usepackage{url}
\usepackage{booktabs}

\begin{document}

\begin{center}
{\Large\bfseries Does Identity-Protective Cognition Operate in Educational Assessment?\\[4pt]
Political Attitudes and Historical Reasoning in Czech Adolescents}

\bigskip

Juda Kaleta\textsuperscript{1}

\medskip

\textsuperscript{1}Institute of History, Faculty of Arts, Charles University, Czech Republic

\medskip

Correspondence: \texttt{juda.kaleta@ff.cuni.cz}

\medskip

ORCID: 0000-0003-0043-4039
\end{center}

\bigskip

\section*{Declarations}

\begin{description}
\item[Competing interests.] The author declares no competing interests.

\item[Funding.] This research received no external funding.

\item[Ethics approval.] This study was conducted in accordance with the ethical requirements of Charles University, Faculty of Arts, and applicable Czech legal provisions for research involving minors. Research procedures were reviewed and approved by the ethics committee of the Institute of History prior to data collection.

\item[Consent to participate.] Participation was fully voluntary; written informed consent was obtained from parents or legal guardians of all participants, and each student's assent was confirmed before administration.

\item[Preregistration.] Hypotheses, instruments, and analysis plan were preregistered on OSF prior to data collection (\url{https://osf.io/yng37/}).

\item[Data availability.] De-identified data, analysis scripts, and preregistration materials are publicly available on the Open Science Framework (\url{https://osf.io/yng37/}). This study shares its dataset with a companion paper reporting the psychometric validation and measurement invariance analysis of the Czech HPT adaptation (Author, in preparation). The two papers address distinct research questions---substantive social psychological questions (present paper) versus measurement properties (companion paper)---and report non-overlapping primary analyses.

\item[Author contribution.] J.K.\ is the sole author and was responsible for all aspects of the study, including conceptualisation, data collection, formal analysis, methodology, writing (original draft), and writing (review and editing).
\end{description}

\end{document}
