%Version 3.1 December 2024
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% Title Page for Double-Blind Peer Review                        %%
%% European Journal of Psychology of Education (EJPE)             %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[pdflatex,sn-apa]{sn-jnl}% APA Reference Style

%%%% Standard Packages
\usepackage{graphicx}%
\usepackage[utf8]{inputenc}%
\usepackage[T1]{fontenc}%
\usepackage{hyperref}

\begin{document}

\title[Political Attitudes and Historical Reasoning]{Do Political Attitudes Contaminate Historical Reasoning? Ideological Attitudes and Disciplinary Thinking in Czech Adolescents}

\author*[1,2]{\fnm{Juda} \sur{Kaleta}}\email{juda.kaleta@ff.cuni.cz}

\affil*[1]{\orgdiv{Institute of History, Faculty of Arts}, \orgname{Charles University}, \orgaddress{\country{Czech Republic}}}

\affil[2]{\orgdiv{Laboratorio Temporal, Faculty of Education}, \orgname{Universidad de Murcia}, \orgaddress{\country{Spain}}}

\maketitle

\section*{Acknowledgments}

The author thanks the participating schools, teachers, and students for making this research possible.

\section*{Declarations}

\subsection*{Competing Interests}

The author declares no competing interests.

\subsection*{Funding}

This research received no external funding.

\subsection*{Ethics Approval}

This study was conducted in accordance with the ethical requirements of Charles University, Faculty of Arts, and applicable Czech legal provisions for research involving minors. Research procedures were reviewed and approved by the ethics committee of the Institute of History prior to data collection.

\subsection*{Consent to Participate}

Participation was fully voluntary; written informed consent was obtained from parents or legal guardians of all participants, and each student's assent was confirmed before administration.

\subsection*{Data Availability}

De-identified data and analysis scripts are available on the Open Science Framework (\url{https://osf.io/yng37/}).

\subsection*{Code Availability}

All analysis code is available on the Open Science Framework (\url{https://osf.io/yng37/}).

\subsection*{Author Contribution}

J.K.\ is the sole author and was responsible for all aspects of the study, including conceptualization, methodology, data collection, formal analysis, visualization, and writing.

\subsection*{Preregistration}

Hypotheses, instruments, and analysis plan were preregistered on the Open Science Framework prior to data collection (\url{https://osf.io/yng37/}).

\end{document}
