% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
  \setmainfont[]{TeX Gyre Pagella}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{lineno}\linenumbers
\usepackage{longtable}
\usepackage{booktabs}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Cross-Cultural Validation and Ideological Fairness of a Historical Perspective Taking Instrument: Evidence from Czech Secondary Students},
  pdfauthor={Juda Kaleta},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Cross-Cultural Validation and Ideological Fairness of a Historical Perspective Taking Instrument: Evidence from Czech Secondary Students}
\author{Juda Kaleta}
\date{2026}

\begin{document}
\maketitle

\setstretch{2}
Juda Kaleta

Institute of History, Faculty of Arts, Charles University, Czech Republic

Email: juda.kaleta@ff.cuni.cz \textbar{} ORCID: 0000-0003-0043-4039

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

This study reports the first Czech adaptation of the Hartmann and Hasselhorn Historical Perspective Taking (HPT) instrument and examines whether its politically charged scenario introduces construct-irrelevant variance related to students' ideological attitudes. Among 293 Czech secondary students, confirmatory factor analysis (CFA) confirmed the three-factor structure. Item Response Theory (IRT)-based differential item functioning (DIF) analysis detected no biased items across ideology groups, and multi-group CFA demonstrated full scalar measurement invariance (meaning the instrument measures the same construct equivalently regardless of students' political attitudes). Historical knowledge predicted HPT scores; ideology was unrelated. These findings support a transferable protocol for evaluating measurement fairness when assessment content carries attitudinal valence, and indicate that structured scenario-based instruments can employ politically sensitive content without compromising score interpretation.

\textbf{Keywords:} historical perspective taking, cross-cultural validation, measurement invariance, differential item functioning, construct-irrelevant variance, test fairness

\hypertarget{introduction}{%
\subsection{1. Introduction}\label{introduction}}

Cross-cultural adaptation of educational and psychological instruments requires more than linguistic translation: it demands evidence that the adapted version measures the same construct, with equivalent psychometric properties, in the target population (American Educational Research Association et al., 2014; Hambleton et al., 2005; International Test Commission, 2017). When assessment content carries attitudinal valence---when scenarios, stimuli, or items touch on beliefs that respondents may hold---an additional fairness question arises: does the content introduce construct-irrelevant variance (CIV) by functioning differently for subgroups whose attitudes align with that content (Messick, 1995)? This question is especially pressing for scenario-based instruments in the social sciences, where rich contextual material is needed to elicit complex reasoning but may simultaneously activate respondents' pre-existing attitudes. Historical perspective taking (HPT)---the capacity to reconstruct the beliefs, values, and circumstances that made past actors' decisions intelligible in their own time (Hartmann \& Hasselhorn, 2008; Lee \& Ashby, 2001)---provides a particularly informative case, because HPT instruments routinely use politically and morally charged historical scenarios as the stimulus for disciplinary reasoning (Drie \& Boxtel, 2007; Seixas \& Morton, 2013; Wineburg, 2001).

The instrument developed by Hartmann \& Hasselhorn (2008) is among the most extensively validated standardized HPT measures in European research. Originally validated with German secondary students, it has been adapted for use in the Netherlands (Huijgen et al., 2014; Huijgen et al., 2017) and employed in comparative studies of historical reasoning. The instrument operationalizes HPT through a scenario-based approach in which students read about a historically situated dilemma and respond to nine Likert-scaled items capturing three developmental levels: presentism (POP), recognition of the agent's role (ROA), and contextualization (CONT). This structure reflects a theoretical progression from naive present-centered judgment toward increasingly sophisticated historical reasoning (Hartmann \& Hasselhorn, 2008; Lee \& Ashby, 2001).

The present study reports the first Czech adaptation and psychometric evaluation of the Hartmann and Hasselhorn HPT instrument. Two considerations motivate this work. First, no validated Czech-language HPT instrument currently exists, despite growing interest in historical thinking assessment within Czech education reform (Činátl et al., 2021; Münich et al., 2023; Ripka et al., 2024). The Czech adaptation addresses this gap and provides the research community with a tool for cross-cultural comparisons of HPT development.

Second, the instrument's scenario content raises a specific fairness question. The scenario describes a young man in the Weimar Republic deliberating whether to join an extremist political movement. Items that require students to contextualize this figure's reasoning could, in principle, function differently for students whose political attitudes align with the scenario's ideological content. Students with right-authoritarian sympathies might find it cognitively easier to endorse contextualizing responses---not because they are reasoning historically, but because the responses resonate with their own attitudes. If present, such a pattern would constitute construct-irrelevant variance (CIV): systematic score variation attributable to a factor extraneous to the target construct (American Educational Research Association et al., 2014; Messick, 1995).

This concern is not merely theoretical. Research on identity-protective cognition demonstrates that individuals process information through the lens of pre-existing beliefs (Jost et al., 2003; Kahan et al., 2017). Recent work has documented how ideological content can infiltrate competence judgments in history assessment, whether through teacher scoring bias (Alvén, 2019) or through construct comparability challenges in cross-cultural settings (Badham et al., 2025). The question central to the present study is whether the HPT instrument's psychometric properties---its factor structure, item parameters, and score interpretations---hold equivalently across students with different political attitudes, or whether politically charged content introduces measurement bias.

The study contributes to the literature in three ways: (a) it demonstrates a transferable methodological approach for evaluating whether attitudinally charged assessment content introduces construct-irrelevant variance, (b) it provides DIF and measurement invariance evidence for a fairness-relevant grouping variable---political ideology---that has not previously been examined in HPT measurement, and (c) it reports the first published Czech validation of the Hartmann and Hasselhorn HPT instrument, extending the cross-cultural evidence base for this measure. To establish this evidence, we integrated confirmatory factor analysis (CFA), reliability analysis, intraclass correlation estimation, IRT-based differential item functioning (DIF) analysis, and multi-group CFA for measurement invariance testing.

\hypertarget{literature-review}{%
\subsection{2. Literature Review}\label{literature-review}}

\hypertarget{historical-perspective-taking-and-its-measurement}{%
\subsubsection{2.1 Historical Perspective Taking and Its Measurement}\label{historical-perspective-taking-and-its-measurement}}

HPT refers to the ability to understand past actors' decisions by reconstructing the historical context in which they operated (Hartmann \& Hasselhorn, 2008; Lee \& Ashby, 2001). The concept has roots in British traditions that progressively refined the construct from historical empathy (Ashby \& Lee, 1987; Shemilt, 1984) toward rational reconstruction (Lee \& Ashby, 2001) and has been situated within broader frameworks of historical reasoning that encompass evidence use, argumentation, and substantive concept deployment (the application of domain-specific content knowledge about events, causes, and actors; Van Boxtel \& Van Drie (2018); Drie \& Boxtel (2007); VanSledright (2014)). The literature variously labels this capacity ``historical empathy,'' ``perspective taking,'' or ``rational reconstruction''; we follow Hartmann \& Hasselhorn (2008) in adopting the term ``perspective taking'' within the rationalist tradition of Lee \& Ashby (2001), which emphasizes cognitive reconstruction of context rather than affective identification with past actors (see Barton \& Levstik, 2004 for a contrasting view that emphasizes affective engagement alongside cognitive reconstruction). Across these frameworks, HPT is understood as a cognitively demanding, knowledge-dependent competence---not a disposition but a disciplinary achievement requiring both domain-specific knowledge and the capacity to deploy that knowledge in reconstructing past perspectives (Chapman, 2021; VanSledright, 2014; Wineburg, 2001).

Hartmann \& Hasselhorn (2008) operationalized HPT into a developmental model with three levels: \emph{presentism} (POP), where students judge historical actors by contemporary standards; \emph{recognition of the agent's role} (ROA), where students acknowledge the agent's social position without full contextual reconstruction; and \emph{contextualization} (CONT), where students situate decisions within the specific historical conditions that rendered them intelligible. The original validation study (\emph{n} = 170 German students, Grade 10) recovered three levels through principal component analysis (explained variance = 51\%), and students classified at the highest level had significantly better history grades. The instrument was extended by Huijgen et al. (2014) and Huijgen et al. (2017), who confirmed the core theoretical structure while highlighting the difficulty of contextualization for most students.

The difficulty of measuring HPT stems from its inherently interpretive character. Unlike factual recall, HPT cannot be assessed through simple multiple-choice items with single correct answers (Breakstone, 2014; Smith, 2017). Scenario-based instruments address this challenge by presenting contextual information and requiring students to evaluate statements reflecting different levels of historical reasoning (Hartmann \& Hasselhorn, 2008; Seixas et al., 2015). However, this approach introduces a measurement tension: the richer the historical context, the more opportunity for construct-irrelevant factors---including attitudes toward that content---to influence responses.

\hypertarget{cross-cultural-test-adaptation}{%
\subsubsection{2.2 Cross-Cultural Test Adaptation}\label{cross-cultural-test-adaptation}}

The adaptation of educational and psychological instruments across languages and cultures requires systematic procedures to ensure that the adapted version measures the same construct as the original (Hambleton et al., 2005; International Test Commission, 2017). The International Test Commission (ITC) guidelines emphasize that adaptation involves more than translation: it requires attention to construct equivalence, cultural appropriateness, and technical quality of the adapted version (International Test Commission, 2017). For instruments used in history education, cross-cultural adaptation faces the additional challenge that historical content may carry different resonances across national contexts. A scenario about the Weimar Republic, originally designed for German students, intersects differently with Czech historical memory---where the period connects directly to the 1938 Munich Agreement, the cession of the Sudetenland, and the Nazi occupation of 1939--1945.

Cross-cultural validation studies should demonstrate that the adapted instrument preserves the factor structure, reliability, and validity relationships of the original (Hambleton et al., 2005). At minimum, this requires confirming the theorized factor structure through CFA, reporting reliability estimates appropriate for the item characteristics, and providing convergent validity evidence through expected correlations with related constructs. When the instrument will be used to compare groups or predict outcomes, measurement invariance testing becomes essential (American Educational Research Association et al., 2014).

\hypertarget{construct-irrelevant-variance-in-educational-assessment}{%
\subsubsection{2.3 Construct-Irrelevant Variance in Educational Assessment}\label{construct-irrelevant-variance-in-educational-assessment}}

The unified validity framework identifies construct-irrelevant variance as a principal threat to score interpretation (American Educational Research Association et al., 2014; Messick, 1995). CIV occurs when systematic variance in test scores is attributable to factors extraneous to the target construct. When assessment content intersects with respondents' attitudes or social identities, CIV can arise through attitude-consistent responding: students whose beliefs align with test content may find certain response options more natural or appealing, inflating their scores without corresponding gains in the assessed competence (Messick, 1995).

For the HPT instrument specifically, the concern is that the Weimar Republic scenario creates a \emph{congruence pathway}---a mechanism whereby attitudinal similarity to item content lowers the cognitive effort required to endorse a response, inflating scores independently of competence. Consider a contextualization item such as ``His decision was understandable given the political and economic turmoil of the Weimar Republic.'' A student who endorses this statement because she has reconstructed the historical context produces a valid HPT response. But a student who endorses it because the sentiment resonates with her own political sympathies---not because she has engaged in historical reasoning---produces an inflated score. This would represent construct contamination in Messick (1995)'s terms---a systematic source of score variance unrelated to HPT competence. Evidence of internal structure, including measurement invariance across relevant subgroups, constitutes essential validity evidence when scores are used to compare groups or when the construct definition requires demonstrating that extraneous factors do not influence scores (American Educational Research Association et al., 2014).

An alternative theoretical prediction deserves mention. Research on authoritarian cognition suggests that high authoritarianism is associated with reduced integrative complexity---the cognitive capacity to hold and reconcile multiple conflicting perspectives simultaneously (Duckitt, 2001; Suedfeld \& Tetlock, 1977). Because HPT requires precisely this kind of integrative capacity, authoritarian students might perform \emph{worse}, not better. The direction of any bias, if present, was genuinely uncertain. Additionally, dual-process theories of cognition (Evans \& Stanovich, 2013) suggest that structured assessment formats may channel deliberative processing, overriding the automatic attitude-retrieval processes that would produce attitude-consistent responding. If so, the assessment format itself may buffer against attitudinal interference.

\hypertarget{dif-and-measurement-invariance-as-fairness-evidence}{%
\subsubsection{2.4 DIF and Measurement Invariance as Fairness Evidence}\label{dif-and-measurement-invariance-as-fairness-evidence}}

Differential item functioning and measurement invariance testing provide complementary approaches to evaluating measurement fairness (Chen, 2007; Cheung \& Rensvold, 2002; Finch, 2005). DIF analysis examines whether individual items function differently across groups after controlling for the latent trait, identifying specific items that may be biased. Measurement invariance testing through multi-group CFA examines whether the overall measurement model holds across groups at three progressively restrictive levels: \emph{configural invariance} (the same factors exist in both groups), \emph{metric invariance} (items relate equally strongly to the factors in both groups), and \emph{scalar invariance} (the scale origin is identical, so mean scores can be directly compared across groups).

For ordinal items, the weighted least squares mean and variance adjusted (WLSMV) estimator is recommended for CFA (Flora \& Curran, 2004), and threshold-based invariance testing replaces intercept-based approaches. The conventional criteria for evaluating invariance steps use changes in fit indices rather than chi-square difference tests, which are sensitive to sample size: \(\Delta\)CFI \(\leq\) .01 and \(\Delta\)RMSEA \(\leq\) .015 indicate that the more constrained model does not fit meaningfully worse (Chen, 2007; Cheung \& Rensvold, 2002).

In the context of HPT measurement, establishing scalar invariance across ideology groups would mean that the factor structure, loadings, and item thresholds are equivalent for students with different political attitudes---a necessary condition for concluding that observed score differences (or the lack thereof) reflect true construct differences rather than measurement artifacts.

\hypertarget{the-present-study}{%
\subsubsection{2.5 The Present Study}\label{the-present-study}}

This study addresses two interrelated measurement questions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Does the Czech adaptation of the Hartmann and Hasselhorn HPT instrument demonstrate acceptable psychometric properties, including the theorized three-factor structure, adequate reliability, and convergent validity with historical knowledge?
\item
  Does the instrument function fairly across students with different political attitudes, as evidenced by the absence of DIF and by scalar measurement invariance across ideology groups?
\end{enumerate}

We hypothesized that the three-factor structure would be confirmed and that the instrument would demonstrate measurement invariance across ideology groups. As convergent validity evidence, we expected historical knowledge to correlate positively with HPT scores, consistent with the theoretical conceptualization of HPT as a knowledge-dependent competence.

\hypertarget{method}{%
\subsection{3. Method}\label{method}}

\hypertarget{adaptation-process}{%
\subsubsection{3.1 Adaptation Process}\label{adaptation-process}}

The Czech HPT instrument was developed through a systematic translation and adaptation process following ITC guidelines (International Test Commission, 2017). Two bilingual Czech-English experts conducted independent forward translations from the published English version of the instrument. An independent backward translation was produced by a third bilingual expert unfamiliar with the original. Five Czech history educators at the secondary level reviewed the adapted items for conceptual equivalence, age-appropriateness, and curricular alignment. No items required substantive modification; minor wording adjustments were made to two items for naturalness in Czech. Reviewers confirmed that the Weimar Republic scenario content was familiar to Czech secondary students through the national history curriculum, given the direct connections between the interwar period and subsequent Czech historical experience (the Munich Agreement, the cession of the Sudetenland, and the Nazi occupation).

The adaptation preserved the original instrument's structure: the four-point response format (1 = \emph{strongly disagree} to 4 = \emph{strongly agree}), the three-factor item structure (three items per subscale), and the reverse-scoring of POP items were retained unchanged from the validated German version. All deviations from the original were documented transparently in the OSF repository (https://doi.org/10.17605/OSF.IO/YNG37). The scenario narrative was cross-checked against the Czech history curriculum to confirm that all contextual references were accessible to the target population.

No cognitive pilot study (e.g., think-aloud interviews) was conducted prior to main data collection, which constitutes a limitation of the adaptation process (see Section 5.6).

\hypertarget{participants}{%
\subsubsection{3.2 Participants}\label{participants}}

The sample comprised 293 students from 20 classrooms across 10 schools in the Czech Republic, spanning lower-secondary (ISCED 2; \emph{n} = 87, 29.7\%) and upper-secondary (ISCED 3; \emph{n} = 206, 70.3\%) levels. The inclusion criterion was completion of the curricular unit on the interwar period. Schools were drawn from five Czech regions (Praha: 42.7\%; Královéhradecký: 34.8\%; Zlínský: 11.3\%; Jihomoravský: 8.2\%; Ústecký: 3.1\%) and three funding categories (public: 90.8\%; private: 6.5\%; church-affiliated: 2.7\%). Gender composition was 192 male (65.5\%), 88 female (30.0\%), and 13 identifying as another gender (4.4\%). Students' most recent history grade averaged 1.71 (\emph{SD} = 0.92; Czech scale: 1 = excellent, 5 = failing).

Schools were recruited through convenience sampling; the sample over-represents upper-secondary students relative to the national age cohort composition and should not be treated as a probability sample of Czech adolescents. Sample size was determined by an a priori power analysis specified in the preregistered protocol (OSF: https://doi.org/10.17605/OSF.IO/YNG37), targeting 80\% power to detect standardized ideology effects of \(\beta\) \(\geq\) .15 in a two-level multilevel model; the achieved \emph{N} = 285--293 met this target. Within each school, all students present on the day of administration completed the instrument battery during a regular class period (approximately 30--40 minutes). Teachers were present during administration but did not intervene in students' responses.

\hypertarget{instruments}{%
\subsubsection{3.3 Instruments}\label{instruments}}

\hypertarget{historical-perspective-taking-hpt}{%
\paragraph{3.3.1 Historical Perspective Taking (HPT)}\label{historical-perspective-taking-hpt}}

Students completed the Czech adaptation of the Hartmann \& Hasselhorn (2008) scenario-based HPT instrument. They read a narrative about a young man in the Weimar Republic considering joining an extremist political movement, then responded to nine items on a 4-point Likert scale. Three items each tapped presentism (POP1--POP3), recognition of the agent's role (ROA1--ROA3), and contextualization (CONT1--CONT3). POP items capture present-centered moral judgments of the historical figure (e.g., evaluating his decision by contemporary standards); these are reverse-scored (5 \(-\) POP, yielding a reversed range of 1--4) so that higher scores indicate more contextualized reasoning. ROA items reflect awareness of the agent's social position and circumstances without full contextual integration. CONT items represent the most sophisticated level, requiring students to situate the historical figure's deliberation within the specific political, economic, and social conditions of the Weimar Republic. The primary composite was HPT\_CTX6, the mean of reversed POP and CONT items. HPT\_TOT9 (adding ROA) served as a secondary outcome.

\hypertarget{right-authoritarian-ideology-fr-lf-mini}{%
\paragraph{3.3.2 Right-Authoritarian Ideology (FR-LF mini)}\label{right-authoritarian-ideology-fr-lf-mini}}

A six-item short form of the Fragebogen Rechtsextremer Einstellungen---Leipziger Form (Questionnaire on Right-Wing Extremist Attitudes---Leipzig Form; Decker et al. (2013)), comprising acceptance of right-wing dictatorship (RD; 3 items) and National Socialist sympathy/relativization (NS; 3 items) facets, rated on a 5-point scale. The NS facet is theoretically the most direct operationalization of potential content congruence with the Weimar scenario.

\hypertarget{authoritarianism-ksa-3}{%
\paragraph{3.3.3 Authoritarianism (KSA-3)}\label{authoritarianism-ksa-3}}

The nine-item Kurzskala Autoritarismus (Beierlein et al., 2014), measuring authoritarian aggression, submission, and conventionalism on a 5-point scale. This instrument was used for a supplementary measurement invariance check using an alternative ideology operationalization.

\hypertarget{historical-knowledge-kn}{%
\paragraph{3.3.4 Historical Knowledge (KN)}\label{historical-knowledge-kn}}

Six multiple-choice items covering content relevant to the HPT scenario, scored dichotomously (sum score 0--6). This measure served as a convergent validity criterion, given the theoretical expectation that HPT is a knowledge-dependent competence.

\hypertarget{social-desirability-sdr-5}{%
\paragraph{3.3.5 Social Desirability (SDR-5)}\label{social-desirability-sdr-5}}

The five-item short form (Hays et al., 1989), rated on a 5-point scale with items SDR2--SDR4 reverse-coded. Included to evaluate whether HPT scores reflect impression management rather than genuine reasoning.

\hypertarget{analytic-approach}{%
\subsubsection{3.4 Analytic Approach}\label{analytic-approach}}

All analyses were conducted in R (version 4.4.2) using lavaan for CFA, mirt for IRT-based DIF analysis, psych for reliability estimation, and lme4/performance for intraclass correlations.

\textbf{Factor structure.} CFA compared one-factor (general HPT), two-factor (POP+CONT vs.~ROA), and three-factor (POP, ROA, CONT) models using the WLSMV estimator, appropriate for ordinal item responses (Flora \& Curran, 2004). CFA fit indices summarize how well a hypothesized factor structure reproduces the observed correlations among items; multiple indices are reported because each captures a different aspect of misfit. Model evaluation used CFI (\(\geq\) .95 acceptable, \(\geq\) .97 good), TLI (\(\geq\) .95), RMSEA (\(\leq\) .08 acceptable, \(\leq\) .05 good), and SRMR (\(\leq\) .08).

\textbf{Reliability.} Internal consistency was assessed via Cronbach's \(\alpha\) (raw and based on polychoric correlations) and McDonald's \(\omega\), which does not assume tau-equivalence and is generally preferred for congeneric measures (Flora \& Curran, 2004).

\textbf{Intraclass correlations.} ICCs at the school and classroom levels were computed to characterize the multilevel structure of the data and to evaluate whether clustering warranted multilevel approaches.

\textbf{Clustered data handling.} The DIF and MG-CFA analyses treat observations as independent, which is standard practice in applied DIF and measurement invariance research. To evaluate whether ignoring the nested structure introduces meaningful bias, design effects (DEFF) were computed post hoc. A design effect of 1.0 means clustering has no impact on precision; values approaching 2.0 indicate modest inflation of standard errors that can usually be tolerated (Maas \& Hox, 2005). The primary composite (HPT\_CTX6) had DEFF = 1.56, within the conventional \(\leq\) 2.0 threshold. The secondary composite (HPT\_TOT9) had DEFF = 2.16, marginally above this threshold; consequently, inferential claims are anchored primarily to HPT\_CTX6, and HPT\_TOT9 results should be interpreted with this caveat.\footnote{DEFF = 1 + (\(\bar{n}_c\) \(-\) 1) \(\times\) ICC, where \(\bar{n}_c\) is the average cluster size (approximately 14.7 students per school). The TYPE=COMPLEX option in lavaan was not applied because the DIF analysis in mirt does not support clustered standard errors, and applying different corrections to DIF and MG-CFA would make the two sets of results less directly comparable.}

\textbf{Differential item functioning.} Low and High ideology groups were formed using tertile splits on a composite ideology score (mean of \emph{z}-scored FR-LF and KSA-3), with the middle tertile excluded to maximize contrast (\emph{n} \(\approx\) 96 per group). This extreme-group design follows established practice in DIF research (Finch, 2005). A constrained multi-group graded response model (GRM item discrimination and threshold parameters for each item by group are reported in Table S1 in the OSF supplementary materials) using an all-others-as-anchor strategy tested each item for uniform DIF (items are systematically easier or harder for one group at all ability levels) and non-uniform DIF (the group difference reverses depending on ability level) with Bonferroni correction (\(\alpha\) = .01). The all-others-as-anchor approach assumes that the remaining eight items are DIF-free when testing each focal item; iterative purification was not applied, which means that if multiple items exhibited small DIF in the same direction, anchor contamination could mask detection (Woods, 2009).

\textbf{Sensitivity of DIF detection.} With approximately 96 respondents per ideology group, the likelihood-ratio DIF tests at Bonferroni-corrected \(\alpha\) = .01 (9 tests) are powered primarily to detect moderate-to-large parameter differences (\(\Delta\beta\) \(\geq\) 0.5 logits on discrimination or threshold parameters). Weak DIF effects (\(\Delta\beta\) \textless{} 0.5 logits) cannot be confidently excluded. This sensitivity threshold is consistent with the sample size requirements documented for IRT-based DIF methods (Finch, 2005). The extreme-group design maximizes statistical contrast between ideology subgroups at the cost of reduced per-group sample size---a standard trade-off in DIF research when the grouping variable is continuous.

\textbf{Multi-group CFA.} Stepwise MG-CFA tested configural, metric, and scalar invariance across ideology groups using the three-factor model with WLSMV estimation. Invariance was evaluated using \(\Delta\)CFI \(\leq\) .01 and \(\Delta\)RMSEA \(\leq\) .015 criteria (Chen, 2007; Cheung \& Rensvold, 2002).

\textbf{Convergent validity.} Pearson correlations between HPT composites and historical knowledge provided convergent validity evidence.

\textbf{Supplementary invariance check.} To verify that invariance findings were not an artifact of the composite ideology definition, MG-CFA was also conducted using NS-only grouping (the facet most directly content-congruent with the Weimar scenario).

\hypertarget{results}{%
\subsection{4. Results}\label{results}}

\hypertarget{factor-structure}{%
\subsubsection{4.1 Factor Structure}\label{factor-structure}}

CFA supported the three-factor model (POP, ROA, CONT) as best-fitting, outperforming the two-factor and one-factor alternatives (Table 1). The three-factor model showed excellent global fit: CFI = .999, TLI = .998, RMSEA = .011 (90\% CI {[}.000, .051{]}), SRMR = .048. The two-factor model combining POP and CONT items on a single factor showed acceptable but inferior fit (CFI = .956, RMSEA = .057), while the one-factor model fit poorly (CFI = .926, RMSEA = .073).

\textbf{Table 1.} \emph{CFA Fit Indices for HPT Factor Models (WLSMV Estimator)}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1296}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1852}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1111}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0926}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0926}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1296}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1481}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1111}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Model
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\(\chi^2\)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\emph{df}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
CFI
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
TLI
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
RMSEA
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
90\% CI
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
SRMR
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
1-Factor & 66.68 & 27 & .926 & .902 & .073 & {[}.051, .095{]} & .075 \\
2-Factor (POP+CONT, ROA) & 49.48 & 26 & .956 & .940 & .057 & {[}.032, .081{]} & .066 \\
3-Factor (POP, ROA, CONT) & 24.78 & 24 & .999 & .998 & .011 & {[}.000, .051{]} & .048 \\
\end{longtable}

\emph{Note.} WLSMV = weighted least squares mean and variance adjusted. All items treated as ordered categorical. The 2-factor model groups POP and CONT items on a single factor with ROA on a separate factor. Chi-square values are mean-and-variance adjusted (WLSMV). Nested model comparisons require the DIFFTEST procedure rather than standard chi-square difference tests. All \emph{p} \textless{} .001 except the three-factor model (\emph{p} = .418).

The near-perfect global fit of the three-factor model reflects the small item set (nine items in three just-identified three-indicator factors) and should be interpreted alongside local fit diagnostics: the largest modification index was 7.35 (below the threshold of 10), and standardized residual correlations were small (\(|r|\) \(\leq\) .13). With only \emph{df} = 24 for the three-factor model (9 items, 3 factors, 24 free parameters against 45 observed statistics), there is minimal room for misfit; the near-unity CFI is therefore a structural consequence of the low-\emph{df} specification rather than evidence of an unusually well-fitting model. These results confirm that the Czech adaptation preserves the theorized three-level structure of the original German instrument.

Inter-factor correlations from the three-factor CFA were: POP--ROA \(\Phi\) = .377, POP--CONT \(\Phi\) = .509, ROA--CONT \(\Phi\) = .631. These moderate-to-strong correlations indicate meaningful associations among the HPT dimensions, with ROA and CONT showing the strongest link. Discriminant validity is supported since all inter-factor correlations are below .85, confirming that the three subscales capture related but distinct aspects of historical perspective taking.

A supplementary bifactor model (one general factor, three specific factors) fit well (\(\chi^2\) = 14.70, \emph{df} = 18, \emph{p} = .682) but did not meaningfully improve over the correlated three-factor model (\(\Delta\)CFI = .001). CONT items loaded primarily on the general factor (.47--.63), while POP and ROA retained meaningful specific factor variance, supporting subscale-level interpretation.

\hypertarget{standardized-factor-loadings}{%
\subsubsection{4.2 Standardized Factor Loadings}\label{standardized-factor-loadings}}

Table 2 presents the standardized loadings from the three-factor CFA. CONT items loaded most strongly (.61--.72), followed by POP items (.30--.74) and ROA items (.30--.67). POP2 showed the weakest loading (.30), at the conventional minimum threshold for item retention; it was retained to preserve fidelity to the original instrument structure. The range of loadings across subscales reflects the heterogeneity within each three-item set and is consistent with patterns reported in prior validations of short-form instruments.

\textbf{Table 2.} \emph{Standardized Factor Loadings for the Three-Factor HPT Model}

\begin{longtable}[]{@{}lllll@{}}
\toprule\noalign{}
Item & Subscale & Loading & \emph{SE} & 95\% CI \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
POP1 & Presentism & .74 & .099 & {[}.54, .93{]} \\
POP2 & Presentism & .30 & .076 & {[}.15, .44{]} \\
POP3 & Presentism & .50 & .068 & {[}.37, .63{]} \\
ROA1 & Recognition of Agent's Role & .61 & .071 & {[}.47, .75{]} \\
ROA2 & Recognition of Agent's Role & .30 & .074 & {[}.15, .44{]} \\
ROA3 & Recognition of Agent's Role & .67 & .074 & {[}.53, .82{]} \\
CONT1 & Contextualization & .72 & .055 & {[}.61, .82{]} \\
CONT2 & Contextualization & .61 & .059 & {[}.49, .72{]} \\
CONT3 & Contextualization & .64 & .054 & {[}.53, .74{]} \\
\end{longtable}

\emph{Note.} POP items are reverse-scored (higher = more contextualized reasoning). All loadings significant at \emph{p} \textless{} .001. WLSMV estimator with items treated as ordered categorical.

\hypertarget{reliability-and-internal-consistency}{%
\subsubsection{4.3 Reliability and Internal Consistency}\label{reliability-and-internal-consistency}}

Table 3 presents reliability coefficients for each HPT subscale. Raw Cronbach's \(\alpha\) values ranged from .47 (POP) to .64 (CONT); polychoric \(\alpha\) (appropriate for ordinal items) ranged from .53 to .69; McDonald's \(\omega\) ranged from .55 to .69. CONT demonstrated the strongest reliability, consistent with its highest average factor loadings.

\textbf{Table 3.} \emph{Reliability Coefficients for HPT Subscales}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1493}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1045}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1343}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2985}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1343}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1791}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Subscale
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Items
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\(\alpha\) (raw)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\(\alpha\) (polychoric)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\(\omega\) total
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Mean \(r_{ii}\)
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
POP & 3 & .47 & .53 & .55 & .226 \\
ROA & 3 & .50 & .53 & .56 & .249 \\
CONT & 3 & .64 & .69 & .69 & .372 \\
\end{longtable}

\emph{Note.} POP items are reverse-scored. \(\alpha\) = Cronbach's alpha; \(\omega\) = McDonald's omega total. Mean \(r_{ii}\) = mean inter-item correlation.

These values are modest by conventional standards but structurally constrained by the three-item design: with \emph{k} = 3, even moderate inter-item correlations (.20--.35) yield \(\alpha\) in this range. The Spearman-Brown formula (which estimates how reliability would improve if subscale length were doubled) projects \(\alpha\) = .64--.79 for six-item subscales, though this assumes tau-equivalence---an assumption violated when loadings range from .30 to .74 (Table 2)---so the projections should be treated as approximate upper bounds. McDonald's \(\omega\) (.55--.69) is more appropriate given unequal loadings (Table 2). The knowledge--HPT correlations reported below (\(r\) = .22--.34) demonstrate that the instrument carries genuine psychometric signal despite modest subscale reliability. Correcting the ideology--HPT correlation for attenuation yielded a disattenuated \(r\) = -.07, indicating that measurement error does not mask a meaningful ideology association.

\hypertarget{descriptive-statistics}{%
\subsubsection{4.4 Descriptive Statistics}\label{descriptive-statistics}}

Table 4 presents descriptive statistics. HPT scores followed the expected developmental ordering (POP\_rev \textgreater{} ROA \textgreater{} CONT), with contextualization producing the lowest scores (\emph{M} = 2.71). Ideology scores were below the scale midpoint (FR-LF: \emph{M} = 2.48; KSA: \emph{M} = 2.86), reflecting the rarity of strong right-authoritarian endorsement in mainstream adolescent samples.

\textbf{Table 4.} \emph{Descriptive Statistics for Study Variables}

\begin{longtable}[]{@{}llllll@{}}
\toprule\noalign{}
Variable & \emph{n} & \emph{M} & \emph{SD} & Min & Max \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
HPT POP (reversed) & 287 & 2.97 & 0.64 & 1.33 & 4.00 \\
HPT ROA & 287 & 2.80 & 0.68 & 1.00 & 4.00 \\
HPT CONT & 287 & 2.71 & 0.74 & 1.00 & 4.00 \\
HPT CTX6 (primary) & 287 & 2.84 & 0.55 & 1.33 & 4.00 \\
HPT TOT9 & 287 & 2.83 & 0.49 & 1.67 & 3.89 \\
KN Total (0--6) & 293 & 3.04 & 1.62 & 0.00 & 6.00 \\
FR-LF RD & 285 & 2.54 & 0.88 & 1.00 & 5.00 \\
FR-LF NS & 286 & 2.43 & 0.89 & 1.00 & 5.00 \\
FR-LF Total & 284 & 2.48 & 0.75 & 1.00 & 5.00 \\
KSA-3 Total & 283 & 2.86 & 0.62 & 1.00 & 4.60 \\
SDR-5 Total & 283 & 3.02 & 0.63 & 1.00 & 4.60 \\
\end{longtable}

\emph{Note.} HPT items scored on a 1--4 scale; POP items reverse-coded. FR-LF and KSA-3 scored on a 1--5 scale. KN scored as sum of correct responses (0--6).

\hypertarget{intraclass-correlations}{%
\subsubsection{4.5 Intraclass Correlations}\label{intraclass-correlations}}

Intraclass correlations indicated meaningful school-level clustering for most HPT outcomes (HPT\_CTX6: ICC\_school = .041; HPT\_TOT9: ICC\_school = .085; ROA: ICC\_school = .102), consistent with between-school differences in instructional emphasis and student composition. Classroom-level variance was near zero for the primary composite but non-negligible for CONT (ICC\_class = .033) and ROA (ICC\_class = .104). These ICCs justify multilevel modeling for substantive analyses and indicate that the instrument is sensitive to contextual variation in historical reasoning opportunities---a desirable property for a measure intended to capture educationally meaningful differences. The corresponding design effects for the primary composite were modest (HPT\_CTX6: DEFF = 1.56; HPT\_TOT9: DEFF = 2.16; see Section 3.4 for computation and implications).

\hypertarget{dif-analysis}{%
\subsubsection{4.6 DIF Analysis}\label{dif-analysis}}

Table 5 presents DIF results for all nine HPT items. Using a constrained multi-group graded response model with likelihood-ratio tests and Bonferroni-adjusted significance (\(\alpha\) = .01), no items were flagged for either uniform or non-uniform DIF. All adjusted \emph{p}-values exceeded .01, indicating statistically equivalent item parameters across ideology groups. This means that at equal levels of underlying HPT competence, students with low and high ideological attitudes responded to each item in the same way---no individual item functioned differentially as a function of political attitudes.

\textbf{Table 5.} \emph{DIF Results per HPT Item (Likelihood-Ratio Tests from Multi-Group Graded Response Model)}

\begin{longtable}[]{@{}llllll@{}}
\toprule\noalign{}
Item & \(\chi^2\) & \emph{df} & \emph{p} (raw) & \emph{p} (Bonferroni) & Flagged \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
POP1 & 11.18 & 4 & .025 & .222 & No \\
POP2 & 9.42 & 4 & .051 & .462 & No \\
POP3 & 9.50 & 4 & .050 & .448 & No \\
ROA1 & 9.88 & 4 & .043 & .383 & No \\
ROA2 & 2.90 & 4 & .575 & 1.000 & No \\
ROA3 & 8.03 & 4 & .091 & .815 & No \\
CONT1 & 1.65 & 4 & .800 & 1.000 & No \\
CONT2 & 3.45 & 4 & .486 & 1.000 & No \\
CONT3 & 9.43 & 4 & .051 & .461 & No \\
\end{longtable}

\emph{Note.} Likelihood-ratio chi-square statistics from constrained multi-group graded response models. Each test compares a fully constrained baseline (all items equal across groups) to a model freeing discrimination and threshold parameters for the tested item. Low (\emph{n} = 96) and High (\emph{n} = 96) ideology groups formed by tertile split on composite ideology score (middle tertile excluded). Bonferroni-adjusted \(\alpha\) = .01 (9 tests). 0 of 9 items flagged. POP1 showed the largest DIF statistic (\(\chi^2\) = 11.18, \emph{p}\_raw = .025), well above the Bonferroni threshold after correction (\emph{p}\_adj = .222). No item approached significance after correction, indicating statistically equivalent item parameters across ideology groups. GRM item discrimination and threshold parameters are reported in Table S1 in the OSF supplementary materials.

\hypertarget{measurement-invariance}{%
\subsubsection{4.7 Measurement Invariance}\label{measurement-invariance}}

Table 6 presents the stepwise MG-CFA results. The configural model demonstrated adequate fit (CFI = .978, RMSEA = .043, SRMR = .074), confirming that the three-factor structure holds separately in both ideology groups. Constraining factor loadings to equality across groups (metric invariance) produced no meaningful deterioration (\(\Delta\)CFI = .000, \(\Delta\)RMSEA = \(-\).002). Constraining both loadings and thresholds (scalar invariance) similarly maintained fit (\(\Delta\)CFI = +.002, \(\Delta\)RMSEA = \(-\).006). Both transitions satisfy the established criteria of \(|\Delta\)CFI\(|\) \(\leq\) .01 and \(|\Delta\)RMSEA\(|\) \(\leq\) .015 (Chen, 2007; Cheung \& Rensvold, 2002), providing evidence for full scalar invariance.

\textbf{Table 6.} \emph{Multi-Group CFA Fit Indices Across Invariance Levels (WLSMV)}

\begin{longtable}[]{@{}llllll@{}}
\toprule\noalign{}
Model & CFI & RMSEA & SRMR & \(\Delta\)CFI & \(\Delta\)RMSEA \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Configural & .978 & .043 & .074 & --- & --- \\
Metric & .978 & .041 & .081 & .000 & \(-\).002 \\
Scalar & .980 & .035 & .076 & +.002 & \(-\).006 \\
\end{longtable}

\emph{Note.} \(\Delta\)CFI and \(\Delta\)RMSEA computed relative to the preceding model. Invariance criteria: \(|\Delta\)CFI\(|\) \(\leq\) .01, \(|\Delta\)RMSEA\(|\) \(\leq\) .015 (Chen, 2007). Low and High ideology groups formed by tertile split on composite ideology (\emph{n} \(\approx\) 96 per group). Items treated as ordered categorical.

The SRMR increased slightly from configural (.074) to metric (.081) before returning at scalar (.076). This pattern can signal partial loading non-invariance. The largest modification index at the metric step was 11.28 (ROA =\textasciitilde{} POP2 cross-loading in the Low ideology group), marginally above the conventional threshold of 10. This suggests a minor loading difference for POP2 that is consistent with its weak loading (.30) in the overall CFA (Table 2) rather than evidence of systematic group-specific misspecification.

A lavaan warning about negative estimated observed variances occurred for the configural and metric models, which can arise with small per-group samples and WLSMV estimation of ordinal items. The affected parameters were residual variances for ROA items in the Low ideology group. This is a known estimation artifact with three-item factors and does not invalidate the invariance conclusions, but it signals that the per-group sample sizes (\emph{n} \(\approx\) 82--96) are at the lower boundary of what MG-CFA with WLSMV requires (Putnick \& Bornstein, 2016). The invariance results should therefore be treated as convergent evidence alongside the DIF analysis rather than as definitive confirmation of scalar equivalence.

Full scalar invariance means that the factor structure, factor loadings, and item thresholds are equivalent for students with low and high ideological attitudes---a necessary condition for meaningful group comparisons.

\hypertarget{supplementary-invariance-ns-only-grouping}{%
\subsubsection{4.8 Supplementary Invariance: NS-Only Grouping}\label{supplementary-invariance-ns-only-grouping}}

To verify that invariance findings were not an artifact of the composite ideology grouping variable, we repeated the MG-CFA using groups defined by the NS facet alone (National Socialist sympathy/relativization), which represents the most direct theoretical operationalization of content congruence with the Weimar scenario. With the NS tertile split, group sizes were \emph{n} = 126 (Low) and \emph{n} = 125 (High). The supplementary analysis confirmed scalar invariance (configural CFI = .960, RMSEA = .050, SRMR = .070; metric \(\Delta\)CFI = +.002, \(\Delta\)SRMR = +.006; scalar \(\Delta\)CFI = +.014, \(\Delta\)SRMR = \(-\).003), indicating that the instrument functions equivalently even when groups are defined by the ideology facet most proximal to the scenario content. Full fit statistics for this supplementary analysis are available in the OSF repository.

\hypertarget{convergent-validity}{%
\subsubsection{4.9 Convergent Validity}\label{convergent-validity}}

Historical knowledge correlated positively with all HPT outcomes, providing convergent validity evidence consistent with the theoretical expectation that HPT is a knowledge-dependent competence: \emph{r} = .34 (HPT\_CTX6), \emph{r} = .22 (CONT), \emph{r} = .32 (POP\_rev). These moderate correlations indicate that students with stronger knowledge of the relevant historical period scored higher on HPT, while also demonstrating that HPT is not reducible to knowledge alone---the shared variance (approximately 5--12\%) leaves substantial HPT variance attributable to reasoning processes beyond factual recall. A caveat is warranted: the KN items were written to cover content directly relevant to the HPT scenario, so the correlation may partly reflect shared specific-content knowledge rather than the broader ``knowledge-dependent competence'' relationship posited theoretically. A general historical knowledge measure would provide a stronger convergent validity test.

History grade also correlated with HPT\_CTX6 (\emph{r} = \(-\).153, \emph{p} = .010; negative sign reflects the Czech scale where 1 = excellent). By contrast, ideology measures showed negligible correlations with HPT (FR-LF--HPT\_CTX6: \emph{r} = \(-\).05), and social desirability was unrelated (\emph{r} = \(-\).02 to .01). This pattern---significant knowledge correlations, null ideology and social desirability correlations---is consistent with an instrument that measures disciplinary reasoning rather than attitudes or impression management.

\textbf{Table 7.} \emph{Zero-Order Pearson Correlations Among Key Variables}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1154}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1154}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1154}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1731}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1154}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1346}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1154}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1154}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
CTX6
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
CONT
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
POP\_rev
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
KN
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
FR-LF
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
KSA
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
SDR
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
HPT CTX6 & --- & & & & & & \\
HPT CONT & .82*** & --- & & & & & \\
HPT POP\_rev & .76*** & .26*** & --- & & & & \\
KN Total & .34*** & .22*** & .32*** & --- & & & \\
FR-LF Total & \(-\).05 & .01 & \(-\).09 & \(-\).10 & --- & & \\
KSA-3 Total & \(-\).03 & .07 & \(-\).13* & .05 & .52*** & --- & \\
SDR-5 Total & \(-\).02 & .01 & \(-\).05 & .03 & \(-\).19** & \(-\).15* & --- \\
\end{longtable}

\emph{Note.} Pairwise complete observations (\emph{n} = 276--287). HPT POP is reverse-scored. KN = historical knowledge (0--6 sum); FR-LF = right-authoritarian ideology (1--5); KSA-3 = authoritarianism (1--5); SDR-5 = social desirability (1--5). *** \emph{p} \textless{} .001. ** \emph{p} \textless{} .01. * \emph{p} \textless{} .05.

\hypertarget{substantive-context-ideology-hpt-relationships}{%
\subsubsection{4.10 Substantive Context: Ideology-HPT Relationships}\label{substantive-context-ideology-hpt-relationships}}

Ideology was unrelated to HPT scores across all model
specifications: FR-LF \(\beta\) = 0.012, \emph{SE} = 0.067, 95\%
CI [\(-\)0.120, 0.143], \emph{p} = .863 for the primary outcome
(HPT\_CTX6). Historical knowledge was the only consistent
predictor (\(\beta\) = 0.21--0.34, \emph{p} \textless{} .001).
Post hoc equivalence tests used the Two One-Sided Tests (TOST)
procedure, which reverses the logic of null-hypothesis testing by
asking whether an effect is small enough to be declared practically
negligible. The Smallest Effect Size Of Interest (SESOI) was set at
\(\pm\) 0.20 standardized units, corresponding to Cohen's (1988)
threshold for a ``small'' effect. A caveat: the preregistered power
target of \(\beta\) \(\geq\) .15 was narrower than this SESOI,
meaning that effects between .15 and .20 cannot be confidently
excluded. With this boundary, the TOST results confirmed that
ideology effects at or above \(\beta\) = .20 are statistically
ruled out (all TOST \emph{p}s \textless{} .003; 90\% CI for the
primary outcome [\(-\)0.10, 0.12], entirely within the \(\pm\)
0.20 SESOI). The ideology coefficient remained non-significant
across all sensitivity checks, including alternative HPT scoring
composites, different ideology operationalizations, exclusion of
high social desirability responders, random-slope models, and
school fixed-effects specifications (all \emph{p}s \textgreater{}
.43). These substantive findings complement the measurement
evidence: the instrument not only measures equivalently across
ideology groups (DIF, invariance) but also yields scores that are
uncontaminated by political attitudes at the composite level.

\hypertarget{discussion}{%
\subsection{5. Discussion}\label{discussion}}

\hypertarget{summary-of-findings}{%
\subsubsection{5.1 Summary of Findings}\label{summary-of-findings}}

The HPT instrument survived every fairness test. Factor structure replicated, DIF was absent, scalar invariance held, and ideology predicted nothing. This is the first psychometric evaluation of a Czech adaptation of the Hartmann and Hasselhorn HPT instrument and the first investigation of ideological fairness for any HPT measure. The Czech adaptation demonstrated the theorized three-factor structure, acceptable reliability for three-item subscales, meaningful convergent validity with historical knowledge, and sensitivity to school-level variation. Historical knowledge was the only construct that predicted HPT scores; political ideology, authoritarianism, and social desirability were all unrelated.

\hypertarget{factor-structure-confirms-the-three-level-hpt-model}{%
\subsubsection{5.2 Factor Structure Confirms the Three-Level HPT Model}\label{factor-structure-confirms-the-three-level-hpt-model}}

The three-factor model (POP, ROA, CONT) fit substantially better than alternatives, confirming that the Czech adaptation preserves the dimensional structure of the original German instrument. This replication in a linguistically and educationally distinct context confirms that the item grouping holds in a Czech sample (Hambleton et al., 2005; International Test Commission, 2017), though with a near-saturated model the structural confirmation is relatively undemanding (see Section 4.1). CONT items loaded most consistently (.61--.72), while POP2 and ROA2 showed weaker loadings (.30); these were retained for comparability but are candidates for revision in future adaptations.

\hypertarget{addressing-modest-reliability}{%
\subsubsection{5.3 Addressing Modest Reliability}\label{addressing-modest-reliability}}

The subscale reliabilities (\(\alpha\) = .47--.64; \(\omega\) = .55--.69) fall below conventional benchmarks but are structurally constrained by the three-item design---a feature of the original instrument, not a deficiency of the Czech adaptation. With \emph{k} = 3 and inter-item correlations of .22--.37, Spearman-Brown projections yield \(\alpha\) = .64--.79 for six-item subscales. Despite modest reliability, the instrument detected theoretically expected knowledge--HPT relationships (\(r\) = .22--.34) and discriminated between HPT levels in the expected ordering. The disattenuated ideology--HPT correlation (\emph{r} = -.065) confirms that even with perfect reliability, no meaningful ideology association would emerge. We recommend using composite scores (HPT\_CTX6 or HPT\_TOT9) rather than individual subscales for applied purposes.

\hypertarget{measurement-fairness-evidence}{%
\subsubsection{5.4 Measurement Fairness Evidence}\label{measurement-fairness-evidence}}

Full scalar invariance---confirmed through both the composite ideology grouping and the supplementary NS-only analysis---means that the HPT instrument imposes equivalent measurement demands on students regardless of their political attitudes. The DIF analysis complements this at the item level: no item exhibited differential functioning, which is particularly important for the CONT items most theoretically susceptible to ideological contamination. The correlation pattern---positive knowledge--HPT associations (Chapman, 2021; Hartmann \& Hasselhorn, 2008; Wineburg, 2001), null ideology correlations, null social desirability correlations---is consistent with the intended score interpretation (American Educational Research Association et al., 2014; Messick, 1995), though the convergent evidence is bounded by the scenario-proximal nature of the knowledge measure (see Section 4.9).

The present fairness evidence applies to mainstream Czech adolescent samples with restricted ideological range. Users in contexts with greater political polarization should conduct their own DIF and invariance analyses before assuming equivalent measurement.

\hypertarget{implications-for-assessment-practice}{%
\subsubsection{5.5 Implications for Assessment Practice}\label{implications-for-assessment-practice}}

The core practical message is that politically charged assessment content does not automatically introduce construct-irrelevant variance---but this conclusion is bounded by the present sample's restricted ideological range and should not be generalized beyond structured, closed-ended formats without further evidence.

The evaluation protocol used here provides an evidence-based model for evaluating fairness when assessment content intersects with respondents' attitudes. This protocol is not specific to HPT or history education; it applies to any scenario-based assessment where content carries attitudinal valence---moral reasoning instruments using politically divisive dilemmas, science literacy assessments touching evolution or climate change, and civic education assessments involving controversial policies. The general approach involves five steps:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Identify the attitudinal dimension theoretically most likely to produce CIV.
\item
  Form extreme groups on that dimension.
\item
  Test DIF at the item level.
\item
  Test measurement invariance at the model level.
\item
  Examine convergent and discriminant correlations to verify that construct scores relate to theoretically expected variables and not to the attitudinal dimension.
\end{enumerate}

The format itself may be a fairness mechanism. If scenario-based formats channel deliberative processing and override automatic attitude-retrieval (Evans \& Stanovich, 2013), then the instrument format---not just the construct---buffers against motivated reasoning. However, this hypothesis has an important implication: if the format itself suppresses attitudinal responding, then the present fairness conclusion is conditional on the closed-ended Likert format and cannot be extended to open-ended HPT assessments where ideology could operate through different cognitive pathways.

Several design features may explain the instrument's resistance to attitudinal contamination: the scenario provides contextual information that channels deliberative processing, the Likert items require evaluating propositions about a historical figure rather than expressing personal attitudes, and the school assessment context may activate an academic frame. These are hypotheses, not conclusions---the present study was not designed to test them experimentally, and doing so would require comparing structured and open-ended formats using the same content.

\hypertarget{limitations}{%
\subsubsection{5.6 Limitations}\label{limitations}}

Several limitations qualify these findings. First, the study uses a single country and scenario; whether the fairness findings generalize to other contexts remains to be established. Second, and most consequentially, the ideological range was restricted (only 17.7\% above the midpoint), so the DIF and invariance analyses compare moderately elevated to low ideology, not genuine political extremism. The null ideology finding should be interpreted as ``no evidence of CIV in this range-restricted sample'' rather than as a general demonstration that politically charged content is fair. Third, DIF power was limited: with \emph{n} \(\approx\) 96 per group, the analysis was powered primarily to detect moderate-to-large effects (\(\Delta\beta\) \(\geq\) 0.5 logits); weak DIF effects cannot be confidently excluded. A formal power simulation for the GRM likelihood-ratio test at Bonferroni-corrected \(\alpha\) was not conducted; future work should quantify the minimum detectable DIF magnitude for this design. Fourth, the sample was 65.5\% male. Gender is a documented moderator of authoritarian attitude endorsement in adolescent samples, and no DIF or sensitivity analysis by gender was conducted. If male students drive both ideology variance and the null finding, the fairness conclusion may not hold for more gender-balanced samples. Fifth, modest subscale reliability (\(\alpha\) = .47--.64) constrains individual-level score interpretations; future adaptations should consider expanding subscale length. Omega hierarchical (\(\omega_h\)) from the bifactor model was not computed; this index would clarify what proportion of subscale variance is attributable to the general HPT factor versus specific subscale factors. Sixth, all measures were self-report instruments administered in a single session, and the closed-ended Likert format may not generalize to open-ended HPT assessments where ideology could operate through different mechanisms (Smith, 2017; Smith et al., 2018). Finally, the study lacks response process evidence. The critical unanswered question is whether Czech students actually invoke historical reasoning when responding---or whether the structured format produces contextualizing responses through a different cognitive pathway. Cognitive interview studies could resolve this.

\hypertarget{recommendations}{%
\subsubsection{5.7 Recommendations}\label{recommendations}}

For researchers developing or adapting HPT instruments: (a) report \(\omega\) alongside \(\alpha\) for short subscales, with Spearman-Brown projections for context; (b) evaluate fairness through DIF and measurement invariance for theoretically relevant grouping variables; (c) consider revising weak-loading items (POP2, ROA2) while preserving the theoretical structure; (d) expand subscale length to improve reliability; and (e) replicate fairness evaluations across diverse samples, scenarios, and cultural contexts.

\hypertarget{conclusion}{%
\subsubsection{5.8 Conclusion}\label{conclusion}}

The Czech adaptation of the Hartmann and Hasselhorn HPT instrument preserves the three-factor structure of the original, produces scores that relate to historical knowledge in theoretically expected ways, and functions with measurement equivalence across students with different political attitudes. The instrument measured what it claims to measure---and ideology had nothing to do with the score.

For researchers and assessment developers, the practical implication is that structured scenario-based assessments can employ politically sensitive historical content without compromising fairness---at least in mainstream classroom settings with restricted ideological range. Whether this conclusion extends to more polarised populations, open-ended assessment formats, or different historical scenarios remains an open empirical question. The evaluation protocol demonstrated here---CFA, DIF, MG-CFA, and convergent/discriminant correlations---provides a replicable template for answering that question wherever attitudinally charged content meets competence assessment.

\hypertarget{data-availability-statement}{%
\subsection{Data Availability Statement}\label{data-availability-statement}}

De-identified data, annotated analysis scripts, the complete instrument battery (in Czech), supplementary materials, and all deviations from the original instrument are available on the Open Science Framework (https://doi.org/10.17605/OSF.IO/YNG37). Data are provided in machine-readable formats (RDS and xlsx) accompanied by a full variable codebook (PDF). The analysis pipeline consists of seven numbered R Markdown scripts that reproduce all reported results; a README file documents the directory structure, execution order, software requirements, and mapping of scripts to manuscript tables and figures. Supplementary materials (Table S1: GRM item parameters) are included in the OSF repository. The study design, hypotheses, and analysis plan were preregistered on OSF prior to data collection.

\hypertarget{conflict-of-interest-disclosure}{%
\subsection{Conflict of Interest Disclosure}\label{conflict-of-interest-disclosure}}

The author declares that he complies with the PCI rule of having no financial conflicts of interest in relation to the content of the article, and has no non-financial conflicts of interest.

\hypertarget{ethics-statement}{%
\subsection{Ethics Statement}\label{ethics-statement}}

This study involved secondary school students completing questionnaires during regular class periods. Under Czech law (Act No.~110/2019 Sb., on the processing of personal data), ethics committee approval is not required for non-interventional educational research involving anonymous questionnaires where no personal data are collected. Participation was voluntary; students were informed about the study's purpose and their right to decline participation without consequences. All responses were de-identified at the point of data entry. No personally identifiable information was collected or stored. The study followed the principles of the Declaration of Helsinki for research involving human participants.

\hypertarget{funding}{%
\subsection{Funding}\label{funding}}

This research received no specific grant from any funding agency in the public, commercial, or not-for-profit sectors.

\hypertarget{author-contributions-credit}{%
\subsection{Author Contributions (CRediT)}\label{author-contributions-credit}}

Juda Kaleta: Conceptualization, Methodology, Software, Formal analysis, Investigation, Data curation, Writing --- original draft, Writing --- review \& editing, Visualization, Project administration.

\hypertarget{acknowledgements}{%
\subsection{Acknowledgements}\label{acknowledgements}}

The author thanks the participating schools, teachers, and students for their cooperation. The translation panel members and expert reviewers who contributed to the adaptation process are acknowledged with gratitude.

\hypertarget{references}{%
\subsection{References}\label{references}}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-Alven2019}{}}%
Alvén, F. L. (2019). Bias in teachers' assessments of students' historical narratives. \emph{History Education Research Journal}, \emph{16}(2). \url{https://doi.org/10.18546/HERJ.16.2.10}

\leavevmode\vadjust pre{\hypertarget{ref-AERA2014}{}}%
American Educational Research Association, American Psychological Association, \& National Council on Measurement in Education. (2014). \emph{Standards for educational and psychological testing}. American Educational Research Association. \url{https://www.testingstandards.net/open-access-files.html}

\leavevmode\vadjust pre{\hypertarget{ref-Ashby1987}{}}%
Ashby, R., \& Lee, P. (1987). Children's concepts of empathy and understanding in history. In \emph{The history curriculum for teachers} (pp. 62--88). Falmer Press.

\leavevmode\vadjust pre{\hypertarget{ref-Badham2025}{}}%
Badham, L., Meadows, M., \& Baird, J.-A. (2025). Construct comparability and the limits of post hoc modeling: Insights from {International Baccalaureate} multi-language assessments. \emph{Frontiers in Education}, \emph{10}, 1616879. \url{https://doi.org/10.3389/feduc.2025.1616879}

\leavevmode\vadjust pre{\hypertarget{ref-Barton2004}{}}%
Barton, K. C., \& Levstik, L. S. (2004). \emph{Teaching history for the common good}. Lawrence Erlbaum Associates. \url{https://doi.org/10.4324/9781410610508}

\leavevmode\vadjust pre{\hypertarget{ref-Beierlein2014}{}}%
Beierlein, C., Asbrock, F., Kauff, M., \& Schmidt, P. (2014). \emph{Die {Kurzskala Autoritarismus} ({KSA-3}): {Ein} {ö}konomisches {Messinstrument} zur {Erfassung} dreier {Subdimensionen} autorit{ä}rer {Einstellungen}} (\{GESIS\} Working Papers 2014/35). {GESIS} -- {Leibniz-Institut} f{ü}r {Sozialwissenschaften}.

\leavevmode\vadjust pre{\hypertarget{ref-Breakstone2014}{}}%
Breakstone, J. (2014). Try, try, try again: The process of designing new history assessments. \emph{Theory \& Research in Social Education}, \emph{42}(4), 453--485. \url{https://doi.org/10.1080/00933104.2014.965860}

\leavevmode\vadjust pre{\hypertarget{ref-Chapman2021}{}}%
Chapman, A. (2021). Introduction: Historical knowing and the {``knowledge turn.''} In A. Chapman (Ed.), \emph{Knowing history in schools: Powerful knowledge and the powers of knowledge} (pp. 1--31). UCL Press. \url{https://doi.org/10.14324/111.9781787357303}

\leavevmode\vadjust pre{\hypertarget{ref-Chen2007}{}}%
Chen, F. F. (2007). Sensitivity of goodness of fit indexes to lack of measurement invariance. \emph{Structural Equation Modeling: A Multidisciplinary Journal}, \emph{14}(3), 464--504. \url{https://doi.org/10.1080/10705510701301834}

\leavevmode\vadjust pre{\hypertarget{ref-Cheung2002}{}}%
Cheung, G. W., \& Rensvold, R. B. (2002). Evaluating goodness-of-fit indexes for testing measurement invariance. \emph{Structural Equation Modeling: A Multidisciplinary Journal}, \emph{9}(2), 233--255. \url{https://doi.org/10.1207/S15328007SEM0902_5}

\leavevmode\vadjust pre{\hypertarget{ref-Cinatl2021}{}}%
Činátl, K., Najbert, J., \& Ripka, V. (2021). Historick{á} gramotnost v aplikaci {HistoryLab}---{Realistick{ý}} p{ř}{í}stup k osvojov{á}n{í} didaktick{é} teorie d{ě}jepisu. \emph{Historie -- Ot{á}zky -- Probl{é}my}, \emph{13}(2), 50--70.

\leavevmode\vadjust pre{\hypertarget{ref-Decker2013}{}}%
Decker, O., Kiess, J., \& Brähler, E. (2013). \emph{Rechtsextremismus der mitte: Eine sozialpsychologische gegenwartsdiagnose}. Psychosozial-Verlag. \url{https://doi.org/10.30820/9783837966367}

\leavevmode\vadjust pre{\hypertarget{ref-vanDrie2007}{}}%
Drie, J. van, \& Boxtel, C. van. (2007). Historical reasoning: Towards a framework for analyzing students' reasoning about the past. \emph{Educational Psychology Review}, \emph{20}(2), 87--110. \url{https://doi.org/10.1007/s10648-007-9056-1}

\leavevmode\vadjust pre{\hypertarget{ref-Duckitt2001}{}}%
Duckitt, J. (2001). A dual-process cognitive-motivational theory of ideology and prejudice. In \emph{Advances in experimental social psychology} (Vol. 33, pp. 41--113). Elsevier. \url{https://doi.org/10.1016/S0065-2601(01)80004-6}

\leavevmode\vadjust pre{\hypertarget{ref-Evans2013}{}}%
Evans, J. St. B. T., \& Stanovich, K. E. (2013). Dual-process theories of higher cognition: Advancing the debate. \emph{Perspectives on Psychological Science}, \emph{8}(3), 223--241. \url{https://doi.org/10.1177/1745691612460685}

\leavevmode\vadjust pre{\hypertarget{ref-Finch2005}{}}%
Finch, H. (2005). The {MIMIC} model as a method for detecting {DIF}: Comparison with {Mantel-Haenszel}, {SIBTEST}, and the {IRT} likelihood ratio. \emph{Applied Psychological Measurement}, \emph{29}(4), 278--295. \url{https://doi.org/10.1177/0146621605275728}

\leavevmode\vadjust pre{\hypertarget{ref-Flora2004}{}}%
Flora, D. B., \& Curran, P. J. (2004). An empirical evaluation of alternative methods of estimation for confirmatory factor analysis with ordinal data. \emph{Psychological Methods}, \emph{9}(4), 466--491. \url{https://doi.org/10.1037/1082-989X.9.4.466}

\leavevmode\vadjust pre{\hypertarget{ref-Hambleton2005}{}}%
Hambleton, R. K., Merenda, P. F., \& Spielberger, C. D. (2005). \emph{Adapting educational and psychological tests for cross-cultural assessment}. Lawrence Erlbaum Associates. \url{https://doi.org/10.4324/9781410611758}

\leavevmode\vadjust pre{\hypertarget{ref-Hartmann2008}{}}%
Hartmann, U., \& Hasselhorn, M. (2008). Historical perspective taking: A standardized measure for an aspect of students' historical thinking. \emph{Learning and Individual Differences}, \emph{18}(2), 264--270. \url{https://doi.org/10.1016/j.lindif.2007.10.002}

\leavevmode\vadjust pre{\hypertarget{ref-Hays1989}{}}%
Hays, R. D., Hayashi, T., \& Stewart, A. L. (1989). A five-item measure of socially desirable response set. \emph{Educational and Psychological Measurement}, \emph{49}(3), 629--636. \url{https://doi.org/10.1177/001316448904900315}

\leavevmode\vadjust pre{\hypertarget{ref-Huijgen2014}{}}%
Huijgen, T., Boxtel, C. van, Grift, W. van de, \& Holthuis, P. (2014). Testing elementary and secondary school students' ability to perform historical perspective taking: The constructing of valid and reliable measure instruments. \emph{European Journal of Psychology of Education}, \emph{29}(4), 653--672. \url{https://doi.org/10.1007/s10212-014-0219-4}

\leavevmode\vadjust pre{\hypertarget{ref-Huijgen2017}{}}%
Huijgen, T., Van Boxtel, C., Van De Grift, W., \& Holthuis, P. (2017). Toward historical perspective taking: Students' reasoning when contextualizing the actions of people in the past. \emph{Theory \& Research in Social Education}, \emph{45}(1), 110--144. \url{https://doi.org/10.1080/00933104.2016.1208597}

\leavevmode\vadjust pre{\hypertarget{ref-ITC2017}{}}%
International Test Commission. (2017). \emph{{ITC} guidelines for translating and adapting tests (second edition)}. International Test Commission. \url{https://www.intestcom.org/files/guideline_test_adaptation_2ed.pdf}

\leavevmode\vadjust pre{\hypertarget{ref-Jost2003}{}}%
Jost, J. T., Glaser, J., Kruglanski, A. W., \& Sulloway, F. J. (2003). Political conservatism as motivated social cognition. \emph{Psychological Bulletin}, \emph{129}(3), 339--375. \url{https://doi.org/10.1037/0033-2909.129.3.339}

\leavevmode\vadjust pre{\hypertarget{ref-Kahan2017}{}}%
Kahan, D. M., Peters, E., Dawson, E. C., \& Slovic, P. (2017). Motivated numeracy and enlightened self-government. \emph{Behavioural Public Policy}, \emph{1}(1), 54--86. \url{https://doi.org/10.1017/bpp.2016.2}

\leavevmode\vadjust pre{\hypertarget{ref-Lee2001}{}}%
Lee, P., \& Ashby, R. (2001). Empathy, perspective taking, and rational understanding. In \emph{Historical empathy and perspective taking in the social studies} (pp. 21--50). Rowman \& Littlefield.

\leavevmode\vadjust pre{\hypertarget{ref-Maas2005}{}}%
Maas, C. J. M., \& Hox, J. J. (2005). Sufficient sample sizes for multilevel modeling. \emph{Methodology}, \emph{1}(3), 86--92. \url{https://doi.org/10.1027/1614-2241.1.3.86}

\leavevmode\vadjust pre{\hypertarget{ref-Messick1995}{}}%
Messick, S. (1995). Validity of psychological assessment: Validation of inferences from persons' responses and performances as scientific inquiry into score meaning. \emph{American Psychologist}, \emph{50}(9), 741--749. \url{https://doi.org/10.1037/0003-066x.50.9.741}

\leavevmode\vadjust pre{\hypertarget{ref-Munich2023}{}}%
Münich, J., Najbert, J., Ondrouchová, M., Pinka, J., Ripka, V., Sedlák, P., Spousta, P., \& Veselá, M. (2023). \emph{Z{á}v{ě}re{č}n{á} zpr{á}va projektu {D{ě}jepis+}: Vyhodnocen{í} pokusn{é}ho ov{ě}{ř}ov{á}n{í} {Prom{ě}na} v{ý}uky d{ě}jin 20. Stol. Na {č}esk{ý}ch z{á}kladn{í}ch {š}kol{á}ch ({{``SINUS CZ''}})}. N{á}rodn{í} pedagogick{ý} institut {Č}esk{é} republiky. \url{https://dejepisplus.npi.cz/wp-content/uploads/2024/01/zaverecna-zprava-d-web.pdf}

\leavevmode\vadjust pre{\hypertarget{ref-Putnick2016}{}}%
Putnick, D. L., \& Bornstein, M. H. (2016). Measurement invariance conventions and reporting: The state of the art and future directions for psychological research. \emph{Developmental Review}, \emph{41}, 71--90. \url{https://doi.org/10.1016/j.dr.2016.06.004}

\leavevmode\vadjust pre{\hypertarget{ref-Ripka2024}{}}%
Ripka, V., Sýkorová, P., Münich, J., \& Chvojka, E. (2024). Epistemic cognition triangulated: {What} can we learn about the theory of epistemic beliefs in history from reassessment of its measurement. In H. Åström Elmersjö \& P. Zanazanian (Eds.), \emph{Teachers and the epistemology of history} (pp. 307--324). Springer Nature Switzerland. \url{https://doi.org/10.1007/978-3-031-58056-7_17}

\leavevmode\vadjust pre{\hypertarget{ref-Seixas2015b}{}}%
Seixas, P., Gibson, L., \& Ercikan, K. (2015). A design process for assessing historical thinking: The case of a one-hour test. In \emph{New directions in assessing historical thinking} (pp. 102--116). Taylor \& Francis.

\leavevmode\vadjust pre{\hypertarget{ref-Seixas2013}{}}%
Seixas, P., \& Morton, T. (2013). \emph{The big six -- historical thinking concepts}. Nelson Education. \url{https://archive.org/embed/bigsixhistorical0000seix}

\leavevmode\vadjust pre{\hypertarget{ref-Shemilt1984}{}}%
Shemilt, D. (1984). Beauty and the philosopher: Empathy in history and classroom. In A. K. Dickinson, P. J. Lee, \& P. J. Rogers (Eds.), \emph{Learning history} (pp. 39--84). Heinemann Educational Books.

\leavevmode\vadjust pre{\hypertarget{ref-Smith2017}{}}%
Smith, M. (2017). Cognitive validity: Can multiple-choice items tap historical thinking processes? \emph{American Educational Research Journal}, \emph{54}(10), 1256--1287. \url{https://doi.org/10.3102/0002831217717949}

\leavevmode\vadjust pre{\hypertarget{ref-Smith2018}{}}%
Smith, M., Breakstone, J., \& Wineburg, S. (2018). History assessments of thinking: A validity study. \emph{Cognition and Instruction}, \emph{37}(1), 118--144. \url{https://doi.org/10.1080/07370008.2018.1499646}

\leavevmode\vadjust pre{\hypertarget{ref-Suedfeld1977}{}}%
Suedfeld, P., \& Tetlock, P. (1977). Integrative complexity of communications in international crises. \emph{Journal of Conflict Resolution}, \emph{21}(1), 169--184. \url{https://doi.org/10.1177/002200277702100108}

\leavevmode\vadjust pre{\hypertarget{ref-VanBoxtel2018}{}}%
Van Boxtel, C., \& Van Drie, J. (2018). Historical reasoning: Conceptualizations and educational applications. In S. A. Metzger \& L. M. Harris (Eds.), \emph{The wiley international handbook of history teaching and learning} (1st ed., pp. 149--176). Wiley. \url{https://doi.org/10.1002/9781119100812.ch6}

\leavevmode\vadjust pre{\hypertarget{ref-VanSledright2014}{}}%
VanSledright, B. A. (2014). \emph{Assessing historical thinking and understanding: Innovative designs for new standards}. Routledge. \url{https://doi.org/10.4324/9781315732879}

\leavevmode\vadjust pre{\hypertarget{ref-Wineburg2001}{}}%
Wineburg, S. (2001). \emph{Historical thinking and other unnatural acts---charting the future of teaching the past}. Temple University Press.

\leavevmode\vadjust pre{\hypertarget{ref-Woods2009}{}}%
Woods, C. M. (2009). Empirical selection of anchors for tests of differential item functioning. \emph{Applied Psychological Measurement}, \emph{33}(1), 42--57. \url{https://doi.org/10.1177/0146621607314044}

\end{CSLReferences}

\newpage

\hypertarget{top-disclosure-table}{%
\section{TOP Disclosure Table}\label{top-disclosure-table}}

\textbf{Cross-Cultural Validation and Ideological Fairness of a Historical Perspective Taking Instrument: Evidence from Czech Secondary Students}

Juda Kaleta

Transparency and Openness Promotion (TOP) Guidelines disclosure following PCI Psychology requirements.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2439}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1707}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.5854}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Standard
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Level
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Description / Location
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Citation & Level 2 & All references include DOIs where available; full bibliography in \texttt{references.bib} \\
Data transparency & Level 3 & De-identified data publicly available on OSF: \url{https://doi.org/10.17605/OSF.IO/YNG37} in both RDS and xlsx formats with a full variable codebook (PDF) \\
Analytic code transparency & Level 3 & All analysis scripts publicly available on OSF: \url{https://doi.org/10.17605/OSF.IO/YNG37} (7 R Markdown pipeline scripts, 4 figure R scripts, 2 supplementary analysis R scripts, 1 Makefile) \\
Research materials transparency & Level 3 & Complete instrument battery (Czech-language versions of HPT, FR-LF, KSA-3, SDR-5, and historical knowledge test) publicly available on OSF: \url{https://doi.org/10.17605/OSF.IO/YNG37} \\
Design transparency & Level 2 & Study design fully described in manuscript Method section (Sections 3.1--3.6); deviations from original instrument documented in OSF repository \\
Preregistration of studies & Level 2 & Study preregistered on OSF prior to data collection: \url{https://doi.org/10.17605/OSF.IO/YNG37} \\
Preregistration of analysis plan & Level 2 & Analysis plan (hypotheses, statistical models, sample size justification) preregistered on OSF: \url{https://doi.org/10.17605/OSF.IO/YNG37} \\
Replication & N/A & This is the first Czech adaptation of the HPT instrument; direct replication is not applicable \\
\end{longtable}

\hypertarget{notes}{%
\subsection{Notes}\label{notes}}

\begin{itemize}
\tightlist
\item
  \textbf{Data format:} The de-identified dataset is provided in two formats: \texttt{.RDS} (native R serialisation) and \texttt{.xlsx} (Excel) to ensure accessibility for users of different software environments.
\item
  \textbf{Code functionality:} All R Markdown scripts render to PDF reports included in the \texttt{outputs/} directory. Scripts can be executed sequentially (01 through 07) or via the included Makefile (\texttt{make\ all}). A reproducibility report (\texttt{07\_reproducibility\_report.Rmd}) documents the exact R version and package versions used.
\item
  \textbf{README documentation:} A detailed \texttt{README.md} in the OSF repository explains the directory structure, execution order, software requirements, and mapping of scripts to manuscript tables and figures.
\end{itemize}

\end{document}
